---
title: "Assignment 2: Spatial Analysis and Visualization"
subtitle: "Healthcare Access and Equity in Pennsylvania"
author: "Your Name Here"
date: today
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
---

## Assignment Overview

**Learning Objectives:**
- Apply spatial operations to answer policy-relevant research questions
- Integrate census demographic data with spatial analysis
- Create publication-quality visualizations and maps
- Work with spatial data from multiple sources
- Communicate findings effectively for policy audiences

---

## Part 1: Healthcare Access for Vulnerable Populations

### Research Question

**Which Pennsylvania counties have the highest proportion of vulnerable populations (elderly + low-income) living far from hospitals?**

Your analysis should identify counties that should be priorities for healthcare investment and policy intervention.

### Required Analysis Steps

Complete the following analysis, documenting each step with code and brief explanations:

#### Step 1: Data Collection

**Your Task:**
```{r}
# Load required packages
#| echo: false
#| message: false
#| warning: false

library(sf)
library(tidyverse)
library(tidycensus)
library(scales)
library(RColorBrewer)
library(here)
library(tigris)

# Load spatial data

hospitals <- st_read(here("labs/lab_2/labs/hospitals.geojson"))
boundaries <- st_read(here("labs/lab_2/labs/Pennsylvania_County_Boundaries/Pennsylvania_County_Boundaries.shp"))
tracts <- suppressMessages(suppressWarnings(tracts(state = "PA", cb = TRUE)))

```

**Questions to answer:**
- How many hospitals are in your dataset?

There are 233 hospitals in my dataset.

- How many census tracts?

There are 3445 census tracts. 

- What coordinate reference system is each dataset in?

```{r}
#| echo: false
#| message: false
#| warning: false

st_crs(hospitals)
st_crs(boundaries)
st_crs(tracts)

```
The hospitals and boundaries data sets are in the WGS 84 CRS, while the tracts data set is in NAD83.


---

#### Step 2: Get Demographic Data 

```{r}
# Get demographic data from ACS

vars_65plus <- c(
  male_65_66   = "B01001_020",
  male_67_69   = "B01001_021",
  male_70_74   = "B01001_022",
  male_75_79   = "B01001_023",
  male_80_84   = "B01001_024",
  male_85plus  = "B01001_025",
  female_65_66  = "B01001_044",
  female_67_69  = "B01001_045",
  female_70_74  = "B01001_046",
  female_75_79  = "B01001_047",
  female_80_84  = "B01001_048",
  female_85plus = "B01001_049"
)

tract_demographics2 <- get_acs(
  geography = "tract",
  variables = c(
    total_pop = "B01003_001",
    median_income = "B19013_001",
    vars_65plus
  ),
  state = "PA",
  year = 2024,
  output = "wide"
)

tract_demographics <- tract_demographics2 %>%
  mutate(
    pop_65_plus = rowSums(
      select(., matches("male_.*E|female_.*E")),
      na.rm = TRUE
    )
  )

tract_demographics <- tract_demographics %>%
  select(GEOID, NAME, total_popE, total_popM, median_incomeE, median_incomeM, pop_65_plus)

# Join to tract boundaries

tracts_proj <- st_transform(tracts,3365)

tract_demographics <- tracts_proj %>%
  left_join(tract_demographics, by = "GEOID")

```

**Questions to answer:**
- What year of ACS data are you using?

I'm using 2024 ACS data for this analysis. 

- How many tracts have missing income data?

```{r}
no_income <- tract_demographics %>%
  filter(is.na(median_incomeE))
```

I identified 69 census tracts that have missing income data. 

- What is the median income across all PA census tracts?

```{r }
#| echo: false
#| message: false
#| warning: false

median <- median(tract_demographics$median_incomeE, na.rm = TRUE)
```

The median income across all of PA's census tracts is $75,844

---

#### Step 3: Define Vulnerable Populations 

Identify census tracts with vulnerable populations based on TWO criteria:
1. Low median household income (choose an appropriate threshold)

Threshold will be households under 30% state median income OR with over 30% of total population being elderly. 

2. Significant elderly population (choose an appropriate threshold)

**Your Task:**
```{r}
# Filter for vulnerable tracts based on your criteria

tract_demographics <- tract_demographics %>%
  mutate(percent_65_plus = pop_65_plus / total_popE)

vulnerable_tracts <- tract_demographics %>%
  filter((median_incomeE <= 0.3*75844) | percent_65_plus >= .30)

```

**Questions to answer:**
- What income threshold did you choose and why?

I chose making less than or equal to 30% of state-wide median income to be the income threshold for vulnerability. This is because that is the HUD definition of "very-low income" and the threshold for qualifying for social assistance such as housing choice vouchers. 

- What elderly population threshold did you choose and why?

I chose the threshold of at least 30% of population being over 65, as after plotting the proportion of residents over 65, it showed that most tracts were clustered between 15 and 25%. This made 30% feel like a reasonable cut-off for counties that skewed older than what is typical. 

- How many tracts meet your vulnerability criteria?

281 tracts meet my criteria. 

- What percentage of PA census tracts are considered vulnerable by your definition?

Approximately 8.1% of tracts are deemed vulnerable. 

---

#### Step 4: Calculate Distance to Hospitals 

First I want to visualize tracts and hospitals. 
```{r}
ggplot() +
  geom_sf(data = hospitals, color = "blue") +
  geom_sf(data = tract_demographics, color="red") +
  theme_void() 
```

**Your Task:**
```{r}
# Transform to appropriate projected CRS

tract_demographics <- st_transform(tract_demographics, crs = 3365)
vulnerable_tracts <- st_transform(vulnerable_tracts, crs = 3365)
hospitals <- st_transform(hospitals, crs = 3365)

# Calculate distance from each tract centroid to nearest hospital

tract_centroid <- st_centroid(vulnerable_tracts)

dist_matrix <- st_distance(tract_centroid, hospitals)

nearest_dist <- apply(dist_matrix, 1, min)

vulnerable_tracts$nearest_hospital_dist_ft <- as.numeric(nearest_dist)

vulnerable_tracts <- vulnerable_tracts %>%
  mutate(nearest_hospital_dist_miles = as.numeric(nearest_hospital_dist_ft / 5280))

```
**Questions to answer:**
- What is the average distance to the nearest hospital for vulnerable tracts?

The average distance to the nearest hospital for vulnerable tracts is 4.7 miles.

- What is the maximum distance?

The maximum distance is 29 miles. 

- How many vulnerable tracts are more than 15 miles from the nearest hospital?

Sixteen census tracts are more than 15 miles away from the nearest hospital.

```{r}
#| echo: false
#| message: false
#| warning: false

far_away <- vulnerable_tracts %>%
  filter(nearest_hospital_dist_miles > 15)

```


---

#### Step 5: Identify Underserved Areas 

Define "underserved" as vulnerable tracts that are more than 15 miles from the nearest hospital.

**Your Task:**
```{r}
# Create underserved variable

vulnerable_tracts <- vulnerable_tracts %>%
  mutate(underserved = ifelse(nearest_hospital_dist_miles > 15, "underserved", "well served"),
         underserved_varb = ifelse(nearest_hospital_dist_miles > 15, 1, 0))




```

**Questions to answer:**
- How many tracts are under served?
- What percentage of vulnerable tracts are under served?
- Does this surprise you? Why or why not?

There are sixteen under-served census tracts. This is approximately 5% of all vulnerable tracts. This suprises me, as I thought that more tracts that are vulnerable would be under-served. However, this might indicate that my selection criteria for under-served tracts biases urban areas where hopsitals are likely to be more densely concentrated. 

---

#### Step 6: Aggregate to County Level


**Your Task:**
```{r}
#| echo: false
#| message: false
#| warning: false



# Spatial join tracts to counties

counties <- st_read(here("labs/lab_2/labs/Pennsylvania_County_Boundaries/Pennsylvania_County_Boundaries.shp"))

# Aggregate statistics by county

counties <- st_transform(counties, crs=3365)


tract_demographics <- tract_demographics %>%
  st_join(counties)

demographcis_centroid <- st_centroid(tract_demographics)

dist_matrix2 <- st_distance(demographcis_centroid, hospitals)

nearest_dist2 <- apply(dist_matrix2, 1, min)

tract_demographics$nearest_hospital_dist_ft <- as.numeric(nearest_dist2)

tract_demographics <- tract_demographics %>%
  mutate(nearest_hospital_dist_miles = as.numeric(nearest_hospital_dist_ft / 5280))


tract_demographics <- tract_demographics %>%
  mutate(vulnerable = ifelse((percent_65_plus >= .30 | median_incomeE <= 0.3*75844), 1, 0),
         underserved = ifelse((nearest_hospital_dist_miles > 15), 1, 0)
  )
  
county_analysis <- tract_demographics %>%
  group_by(COUNTY_NAM) %>%
  summarize(vulnerable_tracts_county = sum(vulnerable, na.rm = TRUE),
            underserved_tracts_county = sum(underserved, na.rm = TRUE),
            per_vul_underserved = sum(vulnerable == 1 & underserved == 1, na.rm = TRUE)/
              sum(vulnerable == 1, na.rm = TRUE),
            dist_hosp_all = (mean(nearest_hospital_dist_miles)),
            dist_hosp_vuln = mean(nearest_hospital_dist_miles[vulnerable == 1], na.rm = TRUE),
            total_pop = sum(total_popE, na.rm = TRUE),
            vulnerable_pop = sum(
              ifelse(vulnerable == 1, total_popE, 0),
              na.rm = TRUE
            )
            ) %>%
  mutate(per_vulnerable = vulnerable_pop/total_pop)
            
vul_far_from_hosp <- county_analysis %>%
  filter(dist_hosp_vuln > 15)


```

**Questions to answer:**
- Which 5 counties have the highest percentage of underserved vulnerable tracts?
- Which counties have the most vulnerable people living far from hospitals?
- Are there any patterns in where underserved counties are located?

Bradford, Cameron, Clarion, Clearfield, and Clinton counties have the highest share of underserved vulnerable tracts, with all of them having 100% of vulnerable tracts marked as undeserved. Counties with the most vulnerable people living far from hospitals are Pike, Clearfield, Clinton, Sullivan, and Wyoming Counties. These are the most populous counties with an average vulnerable tract distance to hospital being over 15 miles. Based on what I've mapped, it appears that most underserved tracts are in central and northern Pennsylvania. This makes sense as they are the furthest from major cities such as Pittsburgh and Philadelphia. 

```{r}
ggplot(county_analysis) + 
  geom_sf(aes(fill = underserved_tracts_county)) +
  scale_fill_viridis_c()+
  theme_minimal() + 
  labs(fill = "Number of Underserved Tracts")
```

---

#### Step 7: Create Summary Table 

**Your Task:**
```{r}
# Create and format priority counties table

library(knitr)

table <- county_analysis %>%
  st_drop_geometry() %>%
  arrange(desc(per_vul_underserved)) %>%
  slice(1:10) %>%
  select(
    COUNTY_NAM, 
    vulnerable_tracts_county,
    underserved_tracts_county,
    per_vul_underserved,
    vulnerable_pop
  ) %>%
  kable(
    digits =2,
    col.names = c(
      "County",
      "Vulnerable Tracts",
      "Underserved Tracts",
      "Percent of Vulnerable Population Underserved by Healthcare",
      "Total Vulnerable Population"
    )
  )

table

```

---

## Part 2: Comprehensive Visualization 

### Map 1: County-Level Choropleth 

Create a choropleth map showing healthcare access challenges at the county level.

**Your Task:**
```{r}
# Create county-level access map

ggplot(county_analysis) + 
  geom_sf(aes(fill = per_vul_underserved),
          color = "black", size = 0.2) +
  geom_sf(data = hospitals, color = "red", size = 1, shape = 21, fill = "red") +
  scale_fill_gradient(
    low = "lightblue",
    high = "darkblue",
    name = "Proportion Vulnerable & Underserved"
  )+
  theme_void() +
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 14),
    plot.caption = element_text(size = 10)
  ) +
  labs(
    title = "Access to Healthcare for Vulnerable Populations",
    subtitle = "Aggregated to PA Counties",
    caption = "Data ACS 2024"
  )


```

---

### Map 2: Detailed Vulnerability Map 

Create a map highlighting underserved vulnerable tracts.

**Your Task:**
```{r}
# Create detailed tract-level map

tract_demographics <- tract_demographics %>%
  mutate(vulnerable_underserved = ifelse(vulnerable == 1 & underserved == 1, 1, 0))

ggplot(tract_demographics) + 
  geom_sf(data = boundaries, color = "black", size = 0.5) +
  scale_fill_manual(
    values = c("0"= "white", "1" = "blue"),
    labels = c("All Else", "Vulnerable and Underserved"),
    name = "Category") + 
  geom_sf(aes(fill = factor(vulnerable_underserved)),
          color = "grey", size = 0.05) +
  geom_sf(data = hospitals, color = "red", size = 0.5, shape = 21, fill = "red") +
  theme_void() +
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 14),
    plot.caption = element_text(size = 10)
  ) +
  labs(
    title = "Tracts that are Vulnerable and Underserved",
    subtitle = "Census Tract Level",
    caption = "Data ACS 2024"
  )
```

---

### Chart: Distribution Analysis

Create a visualization showing the distribution of distances to hospitals for vulnerable populations.

**Your Task:**
```{r}
# Create distribution visualization

ggplot(vulnerable_tracts, aes(x = nearest_hospital_dist_miles)) +
  geom_histogram(binwidth = 2, fill = "steelblue", color = "white", alpha = 0.8) +
  scale_x_continuous(name = "Distance to Nearest Hospital (miles)") +
  scale_y_continuous(name = "Number of Vulnerable Tracts") +
  labs(
    title = "Distribution of Distances to Nearest Hospital",
    subtitle = "Vulnerable Census Tracts in Pennsylvania",
    caption = "Data: ACS 2024"
  ) +
  theme_minimal()


```

The above histogram suggests that the majority of vulnerable census tracts are well-served by hospitals. However, there are extreme outliers that deserve considerable attention, with one vulnerable census tract being 30 miles away from the nearest hospital. 

---

## Part 3: Bring Your Own Data Analysis 

Choose your own additional spatial dataset and conduct a supplementary analysis.

### Investigating Cultural Ammenities in Low-income minority neighborhoods

```{r}
# Load your additional dataset

landmarks <- st_read(here("labs/lab_2/labs/Landmark_Points.geojson"))
#public_art <- st_read(here("labs/lab_2/labs/percent_for_art_public.geojson"))

landmarks <- st_transform(landmarks, crs = st_crs(tract_demographics))
#public_art <- st_transform(public_art, crs = st_crs(tract_demographics))

vars <- vars <- c(
  median_income = "B19013_001", 
  total_pop      = "B02001_001",
  white_pop      = "B02001_002"
)


phila_demos <- get_acs(
  geography = "tract",
  state = "PA",
  county = "Philadelphia",
  variables = vars,
  year = 2024,
  output = "wide",
  geometry = TRUE
) %>%
  mutate(percent_nonwhite = (1-(white_popE / total_popE))*100)

```

I chose the landmarks dataset provided by OpenDataPhilly. I chose these two to get an understanding of the spatial distribution of city landmarks based on socio-economic patterns. The landmarks data set was last updated in 2017 and uses WGS 84. I transformed it to PA state-plane south for better accuracy. 


---


My research will examine whether non-white and low-income census tracts (defined as less than median income) have access to art and cultural installations within approximate walking distance. Walking distance will be defined as a quarter-mile radius from the centroid of the census tract. Furthermore, I will analyze hot-spot patterns of landmarks to determine the demographic characteristics of cultural clusters. 

---


**Your Task:**
```{r}
# Your spatial analysis

#I want to remove neighborhoods from landmarks. 

landmarks <- landmarks %>%
  filter(!SUBTYPE %in% "Neighborhood")

phila_demos <- phila_demos %>%
  mutate(low_inc_minority = ifelse((median_incomeE < median(median_incomeE, na.rm = TRUE)) & (percent_nonwhite > 65), 1, 0)) #creating a category that classifies each tract as low income and minority according to these thresholds. 

phila_demos <- st_transform(phila_demos, st_crs(landmarks))

phila_tracts_centroid <- st_centroid(phila_demos) #finding centroid of each tract.

landmarks_matrix <- st_distance(phila_tracts_centroid, landmarks) #finding distances from centroids to landmarks

nearest_landmark <- apply(landmarks_matrix, 1, min) #finding the closest landmark to each centroid

phila_demos$nearest_landmark <- as.numeric(nearest_landmark) #attaching the nearest landmark distance to my demographic dataset

phila_demos <- phila_demos %>%
  mutate(nearest_landmark_mi = as.numeric(nearest_landmark / 5280)) #converting to miles

phila_demos <- phila_demos %>%
  mutate(low_inc_minority_low_service = ifelse(nearest_landmark_mi < .25 & low_inc_minority == 1, 1, 0))

# Now performing hot-spot analysis of cultural institutions. 

#distance neighbors
#install.packages("spdep")
library(spdep)
library(sp)
 
coords <- st_coordinates(landmarks) #getting coordinates
knn_neighbors <- knearneigh(coords, k = 5) #finding five nearest neighbors of each land mark
neighbors <- knn2nb(knn_neighbors)
weights <- nb2listw(neighbors, style = "W")

#running Getis-Ord Gi*

gi_star <- localG(landmarks$OBJECTID, weights)
landmarks$Gi_star <- as.numeric(gi_star)

#Plotting to see where there are hotspots and cold spots for culture
ggplot() +
  geom_sf(
    data = phila_demos,
    aes(fill = low_inc_minority),
    color = "black",
    size = 0.25
  ) +
  scale_fill_gradient(low = "white", high = "darkgrey", name = "Low Income and Minority Communities") +
  geom_sf(
    data = landmarks %>% filter(!is.na(Gi_star)),
    aes(color = Gi_star),
    size = 2
  ) +
  scale_color_viridis_c(option = "plasma", name = "Gi* Statistic") +
   labs(
    title = "Cultural Clustering and Low-Income Minority Communities",
    subtitle = "Census Tract Level",
    caption = "Data ACS 2024"
  ) +
  theme_void()



```
```{r}
# Now plotting to see where underserved low-income communties are.

ggplot() + 
  geom_sf(
    data = phila_demos,
    fill = "white",      # fill for all tracts
    color = "black", 
    size = 0.25
  ) +
  geom_sf(
    data = phila_demos %>% filter(low_inc_minority_low_service == 1),
    fill = "blue",
    color = "grey",
    size = 0.05
  ) +
  theme_void() +
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 14),
    plot.caption = element_text(size = 10)
  ) +
  labs(
    title = "Low Income Minority Tracts w/o Cultural Ammenities",
    subtitle = "Census Tract Level",
    caption = "Based on 1/4 mile proximity to registered landmarks"
  )
```

```{r}
phila_demos <- phila_demos %>%
  mutate(low_cultural_ammenities = ifelse(nearest_landmark_mi > .25, 1, 0))

table2 <- phila_demos %>%
  st_drop_geometry() %>%
  summarize(
    prop_culturally_underserved = sum(low_inc_minority_low_service, na.rm = TRUE) / sum(low_inc_minority, na.rm = TRUE),
    low_service_prop = mean(low_cultural_ammenities, na.rm = TRUE)
  )

table2 %>%
  kable(
    digits = 2,
    col.names = c(
      "Culturally Underserved Low-Income Minority Population",
      "Culturally Underserved General Population"
    ),
    caption = "Proportion of Tracts within a quarter-mile of a Cultural Ammenity"
  )
```


### Findings

My findings indicate that low-income minority neighborhoods, as defined by over 65% non-white and under Philadelphia median income, are not under served by cultural amenities. In fact, low-income minority neighborhoods are on average closer to cultural amenities, with approximately 65% of census tracts being within a quarter-mile of a registered landmark. However, when looking at the spatial clustering of cultural amenities, hot spots are not clustered in low-income minority communities. This suggests that while low-income and minority communities have access to cultural amenities at a higher rate than the city on average, they do not live in neighborhoods that the land-mark commission deems culturally-rich. 


## Finally - A few comments about your incorporation of feedback!

As recommended in my feedback, I set my set-up code chunks as echo = FALSE so that they are not rendered. I also removed printing to the document to make the user experience cleaner. Finally I removed unnecessary assignment instructions. 

---

